---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Gil Rezin"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

**In this analysis, I look at median age across counties in Washington State, and
ask which communities we measure most and least reliably. Median age is a useful metric to policymakers, as it can be used in determining funding for state programs such as youth or senior programs.**

---

## Choosing what to measure

```{r}
my_state <- "WA"        # two-letter abbreviation, in quotes
my_variable <- "B01002_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

**There are 39 counties in Washington. Therefore, our query returned 39 rows: one per county.**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

**Garfield County has the least reliable estimate, with a margin of error of 8.17%.**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 3  ~ "High confidence",
    moe_pct < 6 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

**Assuming that a high confidence lies with less than 3 percent MOE, 36/39 counties are classified as "High confidence". Of the remaining three, just Garfield County is of Low confidence (>6% MOE). The two remaining counties are of moderate confidence.**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 3  ~ "High confidence",
    moe_pct < 6 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```

**The average population of high-confidence counties is significantly higher than that of low and moderate-confidence counties. Similarly, moderate-confidence counties have higher populations than low-confidence counties. Thus, there appears to exist a direct linear relationship between a county's population and MOE confidence.**

---

## What this means for policy

**As a result of high margins of error, we know the least about the least populous counties in Washington state: Garfield, Wahkiakum, and Columbia counties. Due to its high margin of error, Garfield County has a median age of 52.6 +/- 8.17%. This means that the actual median age of Garfield County is anywhere between 48.3 and 56.9. This is a very wide range, the reason of which comes from the way the ACS works in practice. The 5-year ACS (American Community Survey) is a survey run by the United States Census Bureau that samples approximately 3.5% of the US population per survey conducted. As this is a sample, smaller regions will have less respondents relative to more populous regions. For example, Grant County, WA (population 100,428) would have an approximate sample size of 3,515 people; this yields an MOE of 0.9%. By contrast, a much smaller county such as Columbia County, WA (population 3,996) would only have an approximate sample size of 140 people; this yields in an MOE of 3.6%. When less individuals are sampled overall across a geographic area, the overall result has increased sampling variability, increasing the overall margin of error. This can have downstream effects for policymakers: to not know with accuracy the median age of a county, for example, can result in incorrect funding for schools (hurting younger individuals) or senior support programs (older individuals).**

---

## Optional challenges - For those craving more

---

## Before you submit
