---
title: "Assignment 3: Verify the Claim?"
subtitle: "Testing Axios' claim about citywide change in singlehood"
author: "Gil Rezin"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

According to [reporting by Axios](https://www.axios.com/local/seattle/2025/02/13/single-washington-state-mapped-solo), the share of single-person households in Seattle more than doubled in the course of 5 years, from 17% in 2016 to 40% in 2021. The report utilized ACS (American Community Survey) 1-year data to determine this. Should this statistic be taken at face value?

**In this analysis, I utilize the margins of error provided by the ACS to determine the accuracy of this claim.**

---

## Reconstruct

```{r}
my_state <- "WA"        # two-letter abbreviation, in quotes
my_variables <- c("B11001_008",  "B11001_001")
```

## Pulling the data

```{r}
city_data_t0 <- get_acs(
  geography = "place",
  variables = my_variables,
  state = my_state,
  year = 2016,
  survey = "acs1"
)

city_data_t1 <- get_acs(
  geography = "place",
  variables = my_variables,
  state = my_state,
  year = 2021,
  survey = "acs1"
)
  
single_households_t0 <- city_data_t0[grepl("Seattle", city_data_t0$NAME) & city_data_t0$variable == "B11001_008", ]
total_households_t0 <- city_data_t0[grepl("Seattle", city_data_t0$NAME) & city_data_t0$variable == "B11001_001", ]
single_households_t1 <- city_data_t1[grepl("Seattle", city_data_t1$NAME) & city_data_t1$variable == "B11001_008", ]
total_households_t1 <- city_data_t1[grepl("Seattle", city_data_t1$NAME) & city_data_t1$variable == "B11001_001", ]

single_households_t0_percent <- single_households_t0$estimate / total_households_t0$estimate * 100
single_households_t1_percent <- single_households_t1$estimate / total_households_t1$estimate * 100

cat("2016 -- Single Households / Total Households: ", single_households_t0$estimate, " / ", total_households_t0$estimate, " = ", single_households_t0_percent, "%\n")
cat("2021 -- Single Households / Total Households: ", single_households_t1$estimate, " / ", total_households_t1$estimate, " = ", single_households_t1_percent, "%\n")
```

Already, we have a discrepancy. The article states that in 2021, the rate for single households in 2021 was 40%. This matches my finding. However, in 2016, the article claims a figure of 17%, less than half the rate that I found. 

Let's find our margins of error. Perhaps that could reveal a missing link?

```{r}
# computing percentage rate & its margin of error
single_households_t0_percent_moe <- moe_prop(single_households_t0$estimate, total_households_t0$estimate, single_households_t0$moe, total_households_t0$moe) * 100
single_households_t1_percent_moe <- moe_prop(single_households_t1$estimate, total_households_t1$estimate, single_households_t1$moe, total_households_t1$moe) * 100
cat("2016 margin of error:", single_households_t0_percent_moe, "% ")
cat("2021 margin of error:", single_households_t1_percent_moe, "% ")

single_households <- data_frame(
  year = c(2016, 2021),
  percent_estimate = c(single_households_t0_percent, single_households_t1_percent),
  percent_moe = c(single_households_t0_percent_moe, single_households_t1_percent_moe),
)

# computing coefficient of variation and reliability flag
single_households <- single_households %>%
  mutate(
    cv = (percent_moe / 1.645) / percent_estimate * 100,
    reliability = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
  )
```

```{r}
p_2016 <- single_households %>%
  filter(year == 2016) %>%
  pull(percent_estimate)

p_2021 <- single_households %>%
  filter(year == 2021) %>%
  pull(percent_estimate)

moe_2016 <- single_households %>%
  filter(year == 2016) %>%
  pull(percent_moe)

moe_2021 <- single_households %>%
  filter(year == 2021) %>%
  pull(percent_moe)

se_2016 <- moe_2016 / 1.645
se_2021 <- moe_2021 / 1.645

change <- p_2021 - p_2016
se_diff <- sqrt(se_2016^2 + se_2021^2)
moe_diff <- 1.645 * se_diff

z <- change / se_diff

significant <- abs(z) > 1.645
cat("Is the change in percentage between 2016 and 2021 significant? ", significant)

```
What does this tell us? Not only was there not a doubling in single households in Seattle, as claimed by Axios, there wasn't even any statistical change! This claim is clearly unsupported.

```{r}
library(ggplot2)
library(gt)


df <- data.frame(
  group = c("2016-2021"),
  mean_val = c(change),
  sd_val = c(moe_diff)
)

df_2 <- data.frame(
  group = c("2016", "2021"),
  mean_val = c(p_2016, p_2021),
  sd_val = c(moe_2016, moe_2021)
)

ggplot(df, aes(x = group, y = mean_val)) +
  geom_col(fill = "skyblue", alpha = 0.8, width = 0.6) + # Draw the bars
  geom_errorbar(aes(ymin = mean_val - sd_val, 
                    ymax = mean_val + sd_val), 
                width = 0.2,                             # Width of error bar caps
                color = "grey30",                        # Line color
                linewidth = 0.8) +                       # Thickness of lines
  labs(title = "Percentage Change in Single Households Over Time", 
       x = "Time Period", 
       y = "Percentage Change in single households") +
  theme_minimal()

ggplot(df_2, aes(x = group, y = mean_val)) +
  geom_col(fill = "green3", alpha = 0.8, width = 0.6) + # Draw the bars
  geom_errorbar(aes(ymin = mean_val - sd_val, 
                    ymax = mean_val + sd_val), 
                width = 0.2,                             # Width of error bar caps
                color = "grey30",                        # Line color
                linewidth = 0.8) +                       # Thickness of lines
  labs(title = "Single Households Percentage by Year", 
       x = "Year", 
       y = "Percentage of single households") +
  theme_minimal()


reliability_table <- single_households %>%
  mutate(
    cv = (percent_moe / 1.645) / percent_estimate * 100,
    dot = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    flag = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
  ) %>%
  arrange(desc(cv)) %>%
  select(dot, year, percent_estimate , percent_moe, cv, flag)

reliability_table %>%
  gt() %>%
  fmt_number(columns = c(percent_estimate, percent_moe), decimals = 0) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_label(
    dot      = "",
    year     = "Year",
    percent_estimate = "Percent Estimate",
    percent_moe      = "Percent MOE",
    cv       = "CV (%)",
    flag     = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title    = "Percentage of Single Households",
    subtitle = "City of Seattle, Years 2016 and 2021, ACS 1-year estimates"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )
```


## Adjudication

**Outside of the primary claim examined, the article also makes a few other, relatively easy claims to verify using Census data.**

"Nearly 49% of residents aged 20 and older in the Seattle-Bellevue-Tacoma metro area are single, with 49.4% of women and 48% of men unmarried, per 2019 to 2023 U.S. Census American Community Survey data."
```{r}
metro_areas <- get_acs(
  geography = "cbsa",
  variables = my_variables,
  year = 2023,
  survey = "acs5"
)
percent <- metro_areas[grepl("Seattle", metro_areas$NAME),][2,]$estimate / metro_areas[grepl("Seattle", metro_areas$NAME),][1,]$estimate

cat("Percent of singles: ", percent * 100, "%")
```
Conclusion: Not supported.


"Statewide, 48% of adults are single, compared to the national average of 49%."
```{r}
state_single_rates <- get_acs(
  geography = "state",
  variables = my_variables,
  year = 2023,
  survey = "acs5"
)

percent_statewide <- state_single_rates[grepl("Washington", state_single_rates$NAME),][2,]$estimate / state_single_rates[grepl("Washington", state_single_rates$NAME),][1,]$estimate

cat("Percent of singles: ", percent_statewide * 100, "%")
```
Conclusion: Not Supported.


"Whitman County, home to Washington State University, leads the state in singles, with 60% of residents unmarried."
```{r}
county_single_rates <- get_acs(
  geography = "county",
  variables = my_variables,
  year = 2023,
  state = my_state,
  survey = "acs5"
)

percent_county <- county_single_rates[grepl("Whitman", county_single_rates$NAME),][2,]$estimate / county_single_rates[grepl("Whitman", county_single_rates$NAME),][1,]$estimate

cat("Percent of singles: ", percent_statewide * 100, "%")
```
Conclusion: Not Supported.


## Rewritten central paragraph:
Looking for love? Approximately 40% of the Seattle area's adult residents are unmarried, a trend which has remained consistent over the past several years. Seattle's share of single-person households has remained steady around 40% between 2016 and 2021, according to 1-year American Community Survey data.
